# Part 4: Testing the `mylm` package
```{r, eval = TRUE, echo=FALSE}
library(car)
library(mylm)
data(SLID, package = "carData")
SLID <- SLID[complete.cases(SLID),]
```

In the following section the `mylm` function will be employed to three different model suggestions. To evaluate the fit of these models we'll print the summary of the models to discuss the p-values of the parameters, as well as the residual plot to see if there are any clear trends in the residuals, or discrepancies from the normal assummtion.

\subsection*{Model 1: `sex`, `age`, `language` and `education^2`}

```{r}
model4a <- mylm(formula = wages ~ sex + age + language + I(education^2), data=SLID)
summary(model4a)
```
From the summary of the model fit above the first thing to note is that the intercept is negative, implying that the base-case (all covariates 0) has negative wage. This is non-intuitive, and an indication that the model fit is perhaps not the best.

Further, the p-values for `sex`, `age`and `education^2`indicate that all these parameters have a significant effect on wages, while the effect from `language` is not quite as clear.

Finally, note that the effect from `education` is aggregated, as the effect of wages from an increase in `education` will be quadratic rather than linear. There is no clear indication why this type of response is better than a linear reponse.

```{r}
plot(model4a)
```
In the plot of the residuals a clear trend is visable. From low fitted values the residuals are small, while they increase with larger fitted values. In addition, it seems that the residuals are somewhat asymmetric, with larger positive values than negative. This further support the claim that the model fit is not particularly good.


\subsection*{Model 2: `age`, `language` and interaction term}

```{r}
model4b <- mylm(formula = wages ~ language*age, data=SLID)
summary(model4b)
```
The first thing to notice from the summary print of the second model is that the only two parameters that are \textit{clearly} significant is the intercept term and the effect from `age`. The `language` terms, as well as the interaction effect, is not clearly significant, although the effect of `languageFrench` and `age:languageFrench` would be included on a significance level of 5%. 


```{r}
plot(model4b)
```
The plot of the residuals exhibit a smiliar trend as the previous model, namely that the residuals increase with larger fitted values. In addition, it also seems in this case as if the residuals are asymmetric, the positive residuals are considerably larger in absolute value than the negative. The residuals plot should, if the model was correct, be completeley random. One can with some confidence say that this is not the case for this model.


\subsection*{Model 3: `education` without intercation term}

```{r}
model4c <- mylm(formula = wages ~ -1 + education, data=SLID)
summary(model4c)
```
First thing to notice from this model is that, as there is now intercation term, the only coefficent, `education`, is highly sigificant. The statistics does not make sense in this case, as can be discarded. In all the previous models the interaction term has been significaltly different from 0, indicating that it should be included - there is no indication it should be dropped.

```{r}
plot(model4c)
```
The residual plot seem to have a decreasing trend with the fitted values, as all residuals are larger than zero fro small fitted values, while the majority seem to be negative for larger fitted values. This residual plot does also exhibit asymmetric behaviour, similar to the previous models.
