# Part 2: Simple linear regression with the `mylm` package

### a) Estimation of Coefficients

```{r eval = TRUE}
library(mylm)
model1 <- mylm(wages ~ education, data = SLID)
print(model1)
```

```{r eval = TRUE}
model1b <- lm(wages ~ education, data = SLID)
print(model1b)
```


### b) Estimation of Covariance Matrix


```{r eval = TRUE}
summary(model1)
```


### c) Residual Plot

```{r eval = TRUE}
plot(model1)
```


### d) Residual Analysis

**Q:** What is the residual sum of squares (SSE) and the degrees of freedom for this model? 

**A:** Answer 1.


**Q:** What is total sum of squares (SST) for this model? Test the significance of the regression using a \(\chi^2\)-test. 

**A:** Answer 2.


**Q:** What is the relationship between the \(\chi^2\)- and \(z\)-statistic in simple linear regression? Find the critical value(s) for both tests.

**A:** Answer 3.


### e) Coefficient of Determination

What is the "coefficient of determination" ($R^2$) for this model and how do you interpret this value?
Modify the function `summary.mylm` so that it shows this value.

```{r eval = TRUE}
summary(model1)
```
