# Part 1: Explanatory analysis of the dataset

### Importing the data set

The dataset we will work on is from Canada, and consists of 3987 observations on the following 5 variables:

* `wages`, composite hourly wage rate from all jobs
* `education`, number of years in schooling
* `age`, in years
* `sex`, Male or Female
* `language`, English, French or Other

We will store this data set in a `SLID` dataframe, and remove all rows with incomplete data:

```{r, eval = TRUE}
library(car)
data(SLID, package = "carData")
SLID <- SLID[complete.cases(SLID),]
```

### a) Analyze relationships between some of the variables

First, let's plot a diagnostic plot of all the variables:

```{r, cache = TRUE}
library(GGally)
ggpairs(data = SLID)
```

We can now make some remarks regarding the relationship between `wages` and some of the explanatory variables.

#### Effect of `education` on `wages`

There is correlation of `0.306` between `education` and `wages`, meaning that you would *expect* more educated workers to earn more. 

Looking at the scatter plot, it shows that there is a much greater spread in education among low paid workers, while better wages require a minimum level of education in most cases.

With other words, a high degree of education does not guarantee greater wages, while high wages require higher education.

*PS: Much of the same can be said on the effect of `age` on `wages`.*

#### Effect of `sex` on `wages`

Men earn on average more than women, as low paid jobs are over-represented by women, and high paid jobs are over-represented by men. The extremal values are still approximately equal.
