--- 
title: "TMA4315: Compulsory exercise 3" 
subtitle: "Group 13: Magnus Liland, Jakob Gerhard Martinussen and Emma Skarstein"
date: "`r format(Sys.time(), '%d.%m.%Y')`" # the current date, can be regular text as well
output: 
  # html_document:
  #   toc: yes
  #   toc_depth: 3
  #   toc_float: yes
  pdf_document:
   toc: false
   toc_depth: 2
---

```{r setup, include = FALSE}
library(formatR)
showsol <- FALSE
library(knitr)
opts_chunk$set(tidy.opts = list(width.cutoff = 68), tidy = TRUE, warning = FALSE, error = FALSE, message = FALSE, echo = TRUE)
```

## The dataset

We will use a *simulated* dataset with clustered data. This data is generated from a fitted model to the `jsp` dataset in the `faraway` R-package.

The following variables are made available:

* `school`: 50 schools, with code 1-50.
* `gender`: A factor with levels boy, girl.
* `social`: Social class of the father, categorical.
    Original class 1-2 = S1, 3-4 = S2, 5-6 = S3 and 7-9 = S4
    Note that these are not ordered and S1 is not necessarily higher or lower class than S2!
* `raven`: Test score (centered around 0).
* `math`: Math score (centered around 0).

We will use `math` as response, and group the data by school.

```{r, echo = TRUE}
dataset <- read.table("https://www.math.ntnu.no/emner/TMA4315/2018h/jsp2.txt", header = TRUE)
```

The number of schools is 49, as we omit school number 43 due to the lack of measurements for our particular subset.
